---
title: "Marvel characters clustering"
author: "Sonia Petrini"
date: "6/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(dplyr)
library(StatMatch)
library(parallel)
library(gridExtra)
library(ggplot2)

marvel <- read.csv("marvelcharactersdata.csv")
```

___

# Data preparation

```{r}
summarytools::dfSummary(marvel,graph.col = F,valid.col = F)
```

```{r}
marvel$GSM[marvel$name == "Xavin (Earth-616)"] = "Pansexual Characters"
```

replace empty GSM with heterosexual and change other occurencies names

```{r}
marvel$GSM <- case_when(
  marvel$GSM == "" ~ "heterosexual",
  marvel$GSM == "Bisexual Characters" ~ "bisexual",
  marvel$GSM == "Genderfluid Characters" ~ "genderfluid",
  marvel$GSM == "Homosexual Characters" ~ "homosexual",
  marvel$GSM == "Pansexual Characters" ~ "pansexual",
  marvel$GSM == "Transgender Characters" ~ "transgender",
  marvel$GSM == "Transvestites" ~ "transvestites"
)
```

replace all empty strings with nan and keep only charachters with all information available 

```{r}
marvel[marvel == ""] <- NA
marvel <- na.omit(marvel)
marvel$page_id <-  NULL
marvel$urlslug <-  NULL
```

keep only month of FIRST.APPEARANCE and divide by season

```{r}
library(stringr)
marvel$FIRST.APPEARANCE <- str_sub(marvel$FIRST.APPEARANCE, end=-4)

marvel$FIRST.APPEARANCE <- case_when(
  marvel$FIRST.APPEARANCE %in% c("Jun", "Jul", "Aug") ~ "summer",
  marvel$FIRST.APPEARANCE %in% c("Sep", "Oct", "Nov") ~ "autumn",
  marvel$FIRST.APPEARANCE %in% c("Dec","Jan","Feb") ~ "winter",
  marvel$FIRST.APPEARANCE %in% c("Mar", "Apr", "May") ~ "spring")
```

rename variables values

```{r}
library(stringr)
marvel <- marvel %>%
      mutate_at("name", str_replace, '\\(Earth-616\\)', '') %>%
      mutate_at("ID", str_replace, 'Identity', '') %>%
      mutate_at("ALIGN", str_replace, 'Characters', '') %>%
      mutate_at("EYE", str_replace, 'Eyes', '') %>%
      mutate_at("HAIR", str_replace, 'Hair', '') %>%
      mutate_at("SEX", str_replace, 'Characters', '') %>%
      mutate_at("ALIVE", str_replace, 'Characters', '')
```

set variables to right type

```{r}
marvel[,c(2,3,4,5,6,7,8,10)] <- lapply(marvel[,c(2,3,4,5,6,7,8,10)], as.factor)
```

set the names as the indices

```{r}
rownames(marvel) <- marvel$name
marvel$name <- NULL
```

sample  

```{r}
'%!in%' <- function(x,y)!('%in%'(x,y))

marvel <- marvel %>% filter(ALIVE == "Living ")
sample <- marvel %>% filter(GSM == "heterosexual") %>% slice_head(n=30)
marvel_nohet <- marvel %>% filter(GSM %!in% c("heterosexual","genderfluid"))
marvel <- rbind(marvel_nohet,sample)
marvel$ALIVE <- NULL
rm(sample,marvel_nohet)
```

summary statistic of numerical variables

```{r}
library(xtable)
sum_num <- summary(Filter(is.numeric,marvel))
print(xtable(as.table(sum_num), type = "latex"), file = "sum_num.tex")
```


```{r}
 summarytools::dfSummary(Filter(is.factor ,marvel)[,c(1,2,5,6,7)],graph.col = F,valid.col = F)
```


scale numerical variables

```{r}
s_marvel <- marvel %>% mutate_at(c("APPEARANCES", "Year"), ~(scale(.) %>% as.vector))
```


```{r}
ggplot(marvel, aes(x=Year)) + 
  geom_histogram(color="black", fill="red",binwidth = 7) +
  scale_x_continuous(breaks=c(1941,1951,1961,1971,1981,1991,2001,2011))
ggsave("year_hist.eps",width = 7, height = 3)
```




___

# CLUSTER ANALYSIS

## Hierarchical clustering

```{r}
future::plan(multicore=3)
gower_dist <- gower.dist(s_marvel)
#save(gower_dist, file = "gower_dist_marvel.rda")
```

open the distance matrix in rda

```{r}
#load(file = "gower_dist_marvel.rda")
```

### WARD

```{r}
hc_ward <- hclust(as.dist(gower_dist),method="ward.D2")

setEPS()
#postscript("ward.eps",width = 7, height = 5)
plot(hc_ward, labels = rownames(s_marvel),hang=-5, cex = 0.6)
abline(h=1.5, col="red")
rect.hclust(hc_ward,k=3,border=c("orange","green","blue"))
#dev.off()
```


### 3 clusters
clusters numerosity

```{r}
groups <- cutree(hc_ward, k=3)
s_marvel$clusters <- as.factor(groups)
table(s_marvel$clusters)
```


goodness of clustering

The *Dunn Index* is the ratio of the smallest distance between observations not in the same cluster to the largest intra-cluster distance. The Dunn Index has a value between zero and infinity, and should be maximized.

```{r}
library(clValid)
dunn(distance = as.dist(gower_dist), groups, Data = s_marvel)
```


```{r}
library(factoextra) 

#setEPS()
#postscript("silh.eps",width = 7, height = 4)
viz<-eclust(s_marvel[,-10],"hclust", hc_method = "ward.D2", k=3)
fviz_silhouette(viz) 
#dev.off()

sil<-viz$silinfo$widths 
neg_sil<-which(sil[,'sil_width']<0) 
sil[neg_sil,]
```

### 4 clusters

clusters numerosity

```{r}
groups <- cutree(hc_ward, k=4)
s_marvel$clusters <- as.factor(groups)
table(s_marvel$clusters)
```


goodness of clustering

The *Dunn Index* is the ratio of the smallest distance between observations not in the same cluster to the largest intra-cluster distance. The Dunn Index has a value between zero and infinity, and should be maximized.

```{r}
library(clValid)
dunn(distance = as.dist(gower_dist), groups, Data = s_marvel)
```


```{r}
library(factoextra) 

#setEPS()
#postscript("silh.eps",width = 7, height = 4)
viz<-eclust(s_marvel[,-10],"hclust", hc_method = "ward.D2", k=4)
fviz_silhouette(viz) 
#dev.off()

sil<-viz$silinfo$widths 
neg_sil<-which(sil[,'sil_width']<0) 
sil[neg_sil,]
```


### COMPLETE

```{r}
hc_complete <- hclust(as.dist(gower_dist),method="complete")

setEPS()
#postscript("complete.eps",width = 7, height = 5)
plot(hc_complete, labels = rownames(s_marvel),hang=-5, cex = 0.6)
abline(h=0.8, col="red")
rect.hclust(hc_complete,k=4,border=c("orange","green","blue"))
#dev.off()
```

### 4 clusters
clusters numerosity

```{r}
groups <- cutree(hc_complete, k=4)
s_marvel$clusters <- as.factor(groups)
table(s_marvel$clusters)
```


goodness of clustering

The *Dunn Index* is the ratio of the smallest distance between observations not in the same cluster to the largest intra-cluster distance. The Dunn Index has a value between zero and infinity, and should be maximized.

```{r}
dunn(distance = as.dist(gower_dist), groups, Data = s_marvel)
```


```{r}
#setEPS()
#postscript("silh.eps",width = 7, height = 4)
viz<-eclust(s_marvel[,-10],"hclust", hc_method = "complete", k=4)
fviz_silhouette(viz) 
#dev.off()

sil<-viz$silinfo$widths 
neg_sil<-which(sil[,'sil_width']<0) 
sil[neg_sil,]
```

### 3 clusters

clusters numerosity

```{r}
groups <- cutree(hc_complete, k=3)
s_marvel$clusters <- as.factor(groups)
table(s_marvel$clusters)
```


goodness of clustering

The *Dunn Index* is the ratio of the smallest distance between observations not in the same cluster to the largest intra-cluster distance. The Dunn Index has a value between zero and infinity, and should be maximized.

```{r}
dunn(distance = as.dist(gower_dist), groups, Data = s_marvel)
```


```{r}
#setEPS()
#postscript("silh.eps",width = 7, height = 4)
viz<-eclust(s_marvel[,-10],"hclust", hc_method = "complete", k=3)
fviz_silhouette(viz) 
#dev.off()

sil<-viz$silinfo$widths 
neg_sil<-which(sil[,'sil_width']<0) 
sil[neg_sil,]
```


### AVERAGE

```{r}
hc_average <- hclust(as.dist(gower_dist),method="average")
plot(hc_average, labels = rownames(s_marvel),hang=-5, cex = 0.6)
abline(h=0.6, col="red")
rect.hclust(hc_average,k=3,border=c("orange","green","blue"))
```
The clusters obtained with average linkage look too unbalanced.



```{r}
library(ape)
colors = c("green","blue","orange")
plot(as.phylo(hc_complete), type = "fan", tip.color = colors[s_marvel$clusters],
     label.offset = 0, cex = 0.7)
```


```{r}
library(cluster)

tmp.pr <- cmdscale(gower_dist)
plot(tmp.pr, col=groups)

G <- unique(groups)
for(g in G){
  idx <- which(groups==g)
  if(length(idx)>2){
    xy <- ellipsoidhull(tmp.pr[idx,])
    lines(predict(xy), col=g,lty=2,lwd=2)
  }
}
text(tmp.pr, as.character(rownames(s_marvel)),adj=0, cex = 0.8)
```

___

# CATEGORICAL PCA

## perform PCA

```{r}
library(Gifi)
ordvec <- c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE,FALSE,TRUE)

future::plan(multicore=3)
pca <- princals(s_marvel[,-10], ndim= 4, ordinal =  ordvec)  ## ordinal PCA
summary(pca)
```


### Screeplot

Choose the number of components to retain: 4, explaining 81.7 % of the original variance

```{r}
#setEPS()
#postscript("scree.eps",width = 7, height = 4)
plot(pca, "screeplot")
abline(h=1.0,col="orange",lty=2,lwd=3)
#dev.off()
```


### Components' meaning

* **Component 1**: this component is strongly related to GSM, APPEARANCES, and Year.
  
```{r}
boxplots <- ggplot(marvel, aes(x=reorder(GSM,Year), y=Year, fill=GSM)) +
    theme(legend.position="none") +
    geom_boxplot(varwidth = TRUE) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(x = "Sexual orientation")

dots <- ggplot(marvel, aes(APPEARANCES, Year))+ 
  geom_point(aes(colour = GSM,sixe=3)) + 
  theme(legend.title = element_text(size = 8), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 8)) +
  labs(x = "Number of appearances")

#setEPS()
#postscript("comp1.eps",width = 7, height = 5)
grid.arrange(boxplots,dots,ncol=1)
#dev.off()
```

* **Component 2**: This component is mainly related with ID, HAIR, SEX.
  

```{r}
o <- ggplot(marvel, aes(y=HAIR, x=SEX, colour=HAIR)) +
  geom_count(alpha=0.7) +
  facet_grid(. ~ ID) +
  labs(x = "Type of identity",
       y = "Hair color") +
  scale_size_continuous(range = c(3, 17))+ 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = " ") #+
  #scale_color_manual(values =c("#ffcccc","#3d3936","#ede72d",
                              #"#0066ff","#996600","#cccccc",
                              #"#9966ff","#cc3300","#e67e39",
                              #"#e69b39","#ffffff"))

ggplotly(o)
```

* **Component 3**: This component is mainly related with the season of first appearance, the eye color, but also with identity and hair color.
  
```{r}
ggplot(data=marvel, aes(x=FIRST.APPEARANCE, y=EYE, fill=EYE)) +
  geom_bar(stat="identity") +
  facet_grid(. ~ ID) #+
  #scale_fill_manual(values = c("#38312B","#2496ED","#965C29","#56B85B","grey","#BD9844","#9E3421","#ffccff","white","#EBD642")) + 
  theme(legend.title = element_text(size = 8), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 9)) +
  labs(x = "Season of appearance",
       y = "Eye color")
```

* **Component 4**: This component is the one explaining ALLIGN and EYE.

```{r}
p <- ggplot(marvel, aes(y=EYE, x=ALIGN, colour=ALIGN)) +
  geom_count(alpha=0.4) +
  scale_size_continuous(range = c(1, 23)) + 
  theme(legend.title = element_text(size = 8), 
        legend.text = element_text(size = 9),
        axis.title = element_text(size = 9))
ggplotly(p)
```


```{r}
#setEPS()
#postscript("load.eps",width = 7, height = 4)
plot(pca, "loadplot")
#dev.off()
```


## hierarchical clustering visualization

```{r}
marvel_pca <- data.frame(pca$objectscores)

marvel_pca$clusters <- as.factor(groups)

#setEPS()
#postscript("hier_2d.eps",width = 7, height = 4)
ggplot(marvel_pca, aes(x=D1, y=D2, color=clusters)) + 
  geom_point() + 
  geom_text(label=rownames(marvel_pca)) +
  scale_colour_brewer(
  type = "div",
  palette = "Set2",
  direction = 1,
  aesthetics = "colour")
#dev.off()
```

```{r}
library(threejs)
COLS = RColorBrewer::brewer.pal(3,"Paired")
scatterplot3js(as.matrix(marvel_pca[,c(1,2,3)]),col=COLS[marvel_pca$clusters],size=0.3)
```



___

# K-MEANS WITH PCA

First let's try k-means with 3 clusters, as found in hierarchical methods

```{r}
marvel_kmeans <- marvel_pca %>% select(-clusters)
c_k_means <- kmeans(marvel_kmeans, 3) # k = 3
```

compare with hierarchical clustering groups

```{r}
table(c_k_means$cluster, s_marvel$clusters)
```

*wss plot* for choosing optimal number of clusters

```{r}
wssplot <- function(data, nc=15, seed=1234){
  wss <- (nrow(data)-1)*sum(apply(data,2,var))
  for (i in 2:nc){
    set.seed(seed)
    wss[i] <- sum(kmeans(data, centers=i)$withinss)}
  plot(1:nc, wss, type="b", xlab="Number of Clusters",
       ylab="Within groups sum of squares")
  abline(h=46, col = "orange",lwd=3, lty=2)}

wssplot(marvel_kmeans, nc=6) 
```

Basing on the elbow criteria, we try 3 and 4 clusters

```{r}
marvel_kmeans$clusters <- as.factor(c_k_means$cluster)
ggplot(marvel_kmeans, aes(x=D1, y=D2, color=clusters)) + 
  geom_point() + 
  geom_text(label=rownames(marvel_kmeans))
```

```{r}
library(threejs)

COLS = RColorBrewer::brewer.pal(4,"Set2")
scatterplot3js(as.matrix(marvel_kmeans[,c(1,2,3)]),col=COLS[marvel_kmeans$clusters],size=0.3)

```


The clusters are now more separated and well defined


```{r}
dunn(clusters = as.integer(marvel_kmeans$clusters), Data = marvel_kmeans)
```

```{r}
kmeans_viz<-eclust(marvel_kmeans[,-5],"kmeans", k=3) 
fviz_silhouette(kmeans_viz) 
kmeans_sil<-kmeans_viz$silinfo$widths 
neg_sils<-which(kmeans_sil[,'sil_width']<0) 
kmeans_sil[neg_sils,]
```
### 4 clusters with kmeans

```{r}
c_k_means <- kmeans(marvel_kmeans, 4)
```

```{r}
marvel_kmeans$clusters <- as.factor(c_k_means$cluster)
dunn(clusters = as.integer(marvel_kmeans$clusters), Data = marvel_kmeans)
```

```{r}
kmeans_viz<-eclust(marvel_kmeans[,-5],"kmeans", k=4) 
fviz_silhouette(kmeans_viz) 
kmeans_sil<-kmeans_viz$silinfo$widths 
neg_sils<-which(kmeans_sil[,'sil_width']<0) 
kmeans_sil[neg_sils,]
```

___

## CLUSTERS EXPLORATION (complete, 3 clusters)

```{r}
groups <- cutree(hc_complete, k=3)
marvel$clusters <- as.factor(groups)
id <- marvel %>% 
  group_by(clusters, ID) %>% 
  dplyr::summarise(count = n()) %>% 
  dplyr::mutate(perc = count/sum(count)) %>%
ggplot(aes(x = clusters, y = perc*100, fill = ID)) +
  geom_bar(stat="identity", width = 0.7) +
  labs(x = "Clusters", y = "percentage", fill = "Identity") +
  theme_minimal(base_size = 14) + 
  theme(legend.title = element_text(size = 10), 
               legend.text = element_text(size = 10),
        axis.title = element_text(size = 10)) +
  scale_fill_manual(values=c("#E266E8","#50C7C7","#800B30"))

align <- marvel %>% 
  group_by(clusters, ALIGN) %>% 
  dplyr::summarise(count = n()) %>% 
  dplyr::mutate(perc = count/sum(count)) %>%
ggplot(aes(x = clusters, y = perc*100, fill = ALIGN)) +
  geom_bar(stat="identity", width = 0.7) +
  labs(x = "Clusters", y = "percentage", fill = "Alignment") +
  theme_minimal(base_size = 14) + 
  theme(legend.title = element_text(size = 10), 
               legend.text = element_text(size = 10),
        axis.title = element_text(size = 10)) +
  scale_fill_manual(values=c("#E80049","#56E854","#0FFAFA"))


#setEPS()
#postscript("ID_ALIGN.eps",width = 7, height = 3)
grid.arrange(id,align, ncol=2)
#dev.off()
```

```{r}
eye <- marvel %>% 
  group_by(clusters, EYE) %>% 
  dplyr::summarise(count = n()) %>% 
  dplyr::mutate(perc = count/sum(count)) %>%
ggplot(aes(x = clusters, y = perc*100, fill = EYE)) +
  geom_bar(stat="identity", width = 0.7) +
  labs(x = "Clusters", y = "percentage", fill = "Eye color") +
  theme_minimal(base_size = 14) + 
  theme(legend.title = element_text(size = 10), 
               legend.text = element_text(size = 10),
        axis.title = element_text(size = 10))

hair <- marvel %>% 
  group_by(clusters, HAIR) %>% 
  dplyr::summarise(count = n()) %>% 
  dplyr::mutate(perc = count/sum(count)) %>%
ggplot(aes(x = clusters, y = perc*100, fill = HAIR)) +
  geom_bar(stat="identity", width = 0.7) +
  labs(x = "Clusters", y = "percentage", fill = "Hair color") +
  theme_minimal(base_size = 14) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8),
        axis.title = element_text(size = 10)) #+
  #scale_fill_manual(values =c("#ffcccc","#3d3936","#ede72d",
  #                            "#0066ff","#996600","#cccccc",
  #                            "#9966ff","#cc3300","#e67e39",
  #                            "#e69b39","#ffffff"))
#
#setEPS()
#postscript("EYE_HAIR.eps",width = 7, height = 3.9)
grid.arrange(eye,hair,ncol=2)
#dev.off()
```

```{r}
sex <- marvel %>% 
  group_by(clusters, SEX) %>% 
  dplyr::summarise(count = n()) %>% 
  dplyr::mutate(perc = count/sum(count)) %>%
ggplot(aes(x = clusters, y = perc*100, fill = SEX)) +
  geom_bar(stat="identity", width = 0.7) +
  labs(x = "Clusters", y = "percentage", fill = "Sex") +
  theme_minimal(base_size = 14) + 
  theme(legend.title = element_text(size = 10), 
               legend.text = element_text(size = 10),
        axis.title = element_text(size = 10)) +
  scale_fill_manual(values =c("#ff9999","#ccccff","#00ccff"))

gsm <- marvel %>% 
  group_by(clusters, GSM) %>% 
  dplyr::summarise(count = n()) %>% 
  dplyr::mutate(perc = count/sum(count)) %>%
ggplot(aes(x = clusters, y = perc*100, fill = GSM)) +
  geom_bar(stat="identity", width = 0.7) +
  labs(x = "Clusters", y = "percentage", fill = "Sexual orientation") +
  theme_minimal(base_size = 14) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 10),
        axis.title = element_text(size = 10)) +
  scale_fill_manual(values =c("pink","grey","lightblue","yellow","violet"))

#setEPS()
#postscript("SEX_GSM.eps",width = 7, height = 3)
grid.arrange(sex,gsm,ncol=2)
#dev.off()
```


```{r}
mu_year <- ddply(marvel, "clusters", summarise, grp.mean=mean(Year))

app <- ggplot(marvel, aes(APPEARANCES, clusters, fill = clusters)) +
  geom_violin(trim = FALSE)

year <- ggplot(marvel, aes(x=Year, fill=clusters)) +
  geom_density() +
  geom_vline(data=mu_year, aes(xintercept=grp.mean, color=clusters),
             linetype="dashed") +
  labs(x = "Year of appearance")

first_app <- marvel %>% 
  group_by(clusters, FIRST.APPEARANCE) %>% 
  dplyr::summarise(count = n()) %>% 
  dplyr::mutate(perc = count/sum(count)) %>%
ggplot(aes(x = clusters, y = perc*100, fill = FIRST.APPEARANCE)) +
  geom_bar(stat="identity", width = 0.7) +
  labs(x = "Clusters", y = "percentage", fill = "First appearance") +
  theme_minimal(base_size = 14) + 
  theme(legend.title = element_text(size = 10), 
               legend.text = element_text(size = 10),
        axis.title = element_text(size = 10)) +
  scale_fill_manual(values=c("#ff9933","#99ff66","#0099ff","#99ffff"))

#setEPS()
#postscript("LAST.eps",width = 7, height = 3)
grid.arrange(first_app, arrangeGrob(app,year, ncol=1), ncol=2)
#dev.off()
```








